\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2022-01-22 submission/manuscript.tex
incorporate coauthor feedback

Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn
Flynn\(^1\), Jhansi L Leslie\(^1\), Hamide Sinani\(^1\), Ingrid L.
Bergin\(^3\), Vincent B Young\(^2\), Patrick D. Schloss\(^{1,\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI\\
2. Division of Infectious Diseases, Department of Internal Medicine,
University of Michigan Medical School, Ann Arbor, MI\\
3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
Arbor, MI

@@

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an associate between the gut microbiota disease
severity. The gut microbiota is known to interact with \emph{C.
difficile} during infection. Perturbations to the gut microbiota are
necessary for \emph{C. difficile} to colonize the gut. The gut
microbiota is thought to impair \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
It has not been shown whether the gut microbiota can affect CDI disease
severity. Here we demonstrate that gut bacterial communities can
contribute to disease severity. We derived diverse gut communities by
colonizing germ-free mice with different human fecal communities. The
mice were then infected with a single \emph{C. difficile} ribotype 027
clinical isolate which resulted in moribundity and histopathologic
differences. The variation in severity was associated with the human
fecal community that the mice received. Generally, facultative anaerobes
with pathogenic potential, such as \emph{Escherichia},
\emph{Helicobacter}, and \emph{Klebsiella}, were associated with more
severe outcomes. Bacterial groups associated with fiber degradation,
such as \emph{Coprobacillus}, were associated with less severe outcomes.
Finally, we showed that, when infected with a different \emph{C.
difficile} isolate, the outcome of the infection was similar between the
mice that received the same fecal donor community. These data indicate
that gut bacteria can influence CDI disease severity.

@@

\emph{Clostridioides difficile} infections (CDI) can be asymptomatic or
range in severity from mild diarrhea to severe outcomes including
surgical interventions and death. Models that predict severity and guide
treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has not been investigated for its effect on
CDI disease severity or incorporated into attempts to predict disease
severity. We demonstrated that variation in the microbiome of mice
colonized with human feces yielded a range of disease outcomes. These
results revealed groups of bacteria associated with both severe and mild
\emph{C. difficile} infection outcomes. Gut bacterial community data
from patients with CDI could improve our ability to identify patients at
risk of developing more severe disease and improve interventions which
treat the gut bacteria to reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have increased in
incidence and severity since \emph{C. difficile} was first identified as
the cause of antibiotic-associated pseudomembranous colitis (1). CDI
disease severity can range from asymptomatic to mild diarrhea to toxic
megacolon and death. The Infectious Diseases Society of America (IDSA)
and Society for Healthcare Epidemiology of America (SHEA) guidelines
categorize severe disease based on white blood cell count greater than
15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL,
and complicated disease based on systolic blood pressure below 90,
ileus, or toxic megacolon (2). Since these measures are only present
once the disease has advanced, the IDSA classification does not allow
for identification of patients at risk of severe CDI. Prediction models
have been developed to score a patient's risk for severe CDI outcomes
based on clinical factors but have not been robust for broad application
(3--6). Thus, we have limited ability to prevent patients from
developing severe CDI. Ideally, we could identify patients that are
high-risk of severe disease so their treatments could be adjusted to
halt the infection before the disease worsens.

Missing from CDI severity prediction models are the effects of the
indigenous gut bacteria. \emph{C. difficile} interacts with the gut
community in many ways. The indigenous bacteria of a healthy intestinal
community provide a protective barrier preventing \emph{C. difficile}
from infecting the gut. A range of mechanisms can disrupt this barrier,
including antibiotics, medications, or dietary changes, and lead to
increased susceptibility to CDI (7--9). Once in the gut, the indigenous
bacteria can either promote or inhibit \emph{C. difficile} through
producing molecules or modifying the environment (10, 11). Bile acids
metabolized by the gut bacteria can inhibit \emph{C. difficile} growth
and affect toxin production (12, 13). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (14, 15). While the relationship
between the gut bacteria and \emph{C. difficile} has been established,
the effect the gut bacteria can have on CDI disease severity is unclear.

Studies investigating the severity of CDI disease involving the
microbiome have not shown clear evidence of a relationship between the
microbiome and disease severity. These studies were not setup to
identify a relationship between the microbiome and disease severity.
Studies that have used clinical data cannot control for host, microbiome
or \emph{C. difficile} ribotype variation (16). While disease severity
varies, it cannot be attributed to specifically the microbiome since all
factors vary. Murine experiments typically use a single mouse colony and
different \emph{C. difficile} ribotypes to create severity differences
(3). Recently, our group has begun uncovering the effect microbiome
variation has on \emph{C. difficile} infection. We showed the variation
in the bacterial communities between mice from different mouse colonies
resulted in different clearance rates of \emph{C. difficile} (17). We
also showed varied ability of mice to spontaneously eliminate \emph{C.
difficile} infection when they were treated with different antibiotics
prior to \emph{C. difficile} challenge (18). However, there have not
been any studies investigating whether variation in the microbiome
affects the resultant disease severity of \emph{C. difficile}
infections.

Here, we sought to determine whether the gut bacteria contribute to
disease severity. We hypothesized that since the gut bacteria affect the
metabolism of \emph{C. difficile} and its infection dynamics, it can
also affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal
communities to create varied gut communities. We then challenged the
mice with a single \emph{C. difficile} isolate, ribotype 027 (RT027)
human isolate 431, and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without perturbation.} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (19). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. The human fecal
samples were selected because they represented diverse community
structures based on community clustering (20). The gut communities were
allowed to establish themselves for two weeks post inoculation (21). We
then surveyed the bacterial members of the gut communities by 16S rRNA
gene sequencing of fecal pellets (Figure 1A). The bacterial communities
from each mouse grouped more closely to those communities from mice that
received the same human fecal donor community than to the mice who
received a different human fecal donor community (Figure 1B). The
communities were primarily composed of populations of \emph{Clostridia},
\emph{Bacteroidia}, \emph{Erysipelotrichia}, \emph{Bacilli}, and
\emph{Gammaproteobacteria}. However, the gut bacterial communities of
each donor group of mice harbored unique relative abundance
distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment with antibiotics, such
as clindamycin, to become susceptible to \emph{C. difficile}
colonization (22, 23). However, we wanted to avoid modifying the
communities with an antibiotic to maintain their unique microbial
compositions and ecological relationships. Since these communities came
from people who had increased risk of CDI, such as hospitalization or
recent antibiotic use (20), we decided to test if \emph{C. difficile}
was able to infect these mice without any antibiotic perturbation. We
hypothesized that \emph{C. difficile} would be able to colonize a subset
of mice whose communities were susceptible from perturbations either the
donor experienced or from the ecological selection when colonizing the
murine gut. Mice were challenged with 10\(^{3}\) \emph{C. difficile}
RT027 human isolate 431 spores. The mice were followed for 10 days
post-challenge and their stool was collected and plated for \emph{C.
difficile} colony forming units (CFU) to determine the extent of the
infection. Surprisingly, mice from all donors were colonized (Figure 2).
Only two mice were able to resist \emph{C. difficile} infection and both
received their community from the same donor, N3. By colonizing
germ-free mice with different human fecal communities, we were able to
generate diverse gut communities in mice, which were susceptible to
\emph{C. difficile} infection without further modification of the gut
community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N2 and N5, were not colonized on the day after \emph{C.
difficile} challenge but were infected (\textgreater10\(^{6}\)) by the
end of the experiment. All mice that received their community from Donor
M1 through M6 succumbed to the infection and became moribund within 3
days post-challenge. The remaining mice, except the uninfected Donor N3
mice, maintained \emph{C. difficile} infection through the end of the
experiment (Figure 2). At 10 days post-challenge, or earlier for the
moribund mice, mice were euthanised and fecal pellets were assayed for
toxin activity and cecal tissue was collected and scored for
histopathologic signs of disease (Figure 3). Overall, there was greater
toxin activity detected in the stool of the moribund mice (\emph{P} =
0.003). However, when looking at each group of mice, we observed a range
in toxin activity for both the moribund and non-moribund mice (Figure
3A). Non-moribund mice from Donors N4 through N9 had comparable toxin
activity as the moribund mice. Additionally, not all moribund mice had
toxin activity detected in their stool. This discordance between toxin
activity and disease severity has been previously described (24, 25).
Next we examined the cecal tissue for histopathologic damage. Moribund
mice had high levels of epithelial damage, tissue edema, and
inflammation (Figure S1) similar to previously reported histopathologic
findings for \emph{C. difficile} RT027 (26). As observed with toxin
activity, the moribund mice had higher histopathologic scores than the
non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
toxin activity, all moribund mice had consistently high histopathologic
summary scores (Figure 3B). The non-moribund mice, Donor groups N1
through N9, had a range in tissue damage from none detected to similar
levels as the moribund mice, which grouped by community donor. Together,
the toxin activity, histopathologic score, and moribundity showed
variation across the donor groups but were largely consistent within
each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score. This analysis revealed 20 genera that
were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the observed differences in toxin activity (Figure 4B). Many genera that
associated with toxin production were also associated with moribundity,
such as populations of \emph{Escherichia/Shigella} and
\emph{Bacteroides}. Likewise, there were genera such as
\emph{Anaerotignum}, \emph{Enterocloster}, and \emph{Murimonas} that
were associated with no toxin production that also exhibited greater
relative abundance in communities from non-moribund mice with a low
histopathologic score. Lastly, we tested for correlations between the
endpoint relative abundances of bacterial operational taxonomic units
(OTUs) and the histopathologic summary score (Figure 4C). The endpoint
relative abundance of \emph{Bacteroides} was positively correlated with
histopathologic score, as its day 0 relative abundance did with disease
severity (Figure 4A). Populations of \emph{Klebsiella} and
\emph{Prevotellaceae} were positively correlated with the
histopathologic score and were increased in the group of mice with
detectable toxin. This analysis identified bacterial genera that were
associated with the variation in moribundity, histopathologic score, and
toxin activity of CDI.

@@

\textbf{Disease severity consistent by donor community across isolates.}
Since Carlson et al.~showed \emph{C. difficile} isolates with lower
germination rates were more likely to result in severe disease (27), we
next tested whether the effect of the community on CDI outcomes was
conserved across \emph{C. difficile} isolates with higher rates of
germination. We colonized mice with human fecal slurry of either Donor
N7, N6, or M4 and allowed two weeks for the community to establish. We
then challenged the mice with a different \emph{C. difficile} isolate
and all mice were infected. The moribundity and histopathologic score
outcomes when challenged with a different \emph{C. difficile} human
isolate were comparable to the outcomes of challenges with \emph{C.
difficile} RT027 human isolate 431 by their donor group (Figure 6). For
the isolates and communities we tested, the effect of the gut bacterial
community on disease severity was greater than the effect of the
variation in these isolates.

@@

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
variation in the gut bacteria affects the \emph{C. difficile} infection
disease severity. Our analysis revealed an association between the
relative abundance of bacterial community members and disease severity.
We used the relative abundance patterns of community members associated
with disease outcomes to select donor communities to test whether the
disease severity was conserved when challenged with a different \emph{C.
difficile} isolate. The disease outcomes from the challenge with the
second \emph{C. difficile} isolate matched the outcomes of the challenge
with \emph{C. difficile} RT027 isolate 431. Therefore, the effect of the
community variation on disease severity \emph{in vivo} was greater than
the effect of \emph{C. difficile} isolate variation (27, 28). Overall,
these results have demonstrated that the bacterial community contributed
to the disease severity of \emph{C. difficile} infection.

CDI severity can range from no symptoms to mild symptoms, such as
diarrhea, to severe symptoms, such as toxic megacolon or death.
Physicians use prediction tools to identify patients most at risk of
developing a severe infection using white blood cell counts, serum
albumin level, or serum creatinine level (2, 29, 30). Those levels are
driven by the activities in the intestine (31). Research into the
drivers of this variation have revealed factors that make \emph{C.
difficile} more virulent. Strains are categorized for their virulence by
the presence and production toxin TcdA, toxin TcdB, and binary toxin and
the prevalence in outbreaks, such as ribotypes 027 and 078 (22, 32--35).
However, other studies have shown that virulence is not necessarily
linked with toxin production (36) or the strain (37). Furthermore, there
is variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (27, 28, 38, 39).
This variation may help explain why severe CDI prediction tools often
miss identifying many patients with CDI that will develop severe disease
(3--6). Therefore, it is necessary to gain a full understanding of all
factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy gut community
and is only able to become established after a perturbation (23). Once
colonized, the different communities lead to different metabolic
responses and dynamics of the \emph{C. difficile} population (11, 18,
40). The gut bacteria metabolize primary bile acids into secondary bile
acids (41, 42). The concentration of these bile acids affect the
germination, growth, toxin production and biofilm formation (12, 13, 43,
44). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (45, 46). The nutrient environment
affects toxin production (47). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

Myriad studies have explored the relationship between the microbiome and
CDI but none have established a relationship between the variation of
disease severity and the microbiome. CDI studies often use an infection
model with a single homogeneous bacterial community. Collins et al.~used
multiple human communities to colonize mice, however the communities
were pooled prior to gavaging into germ-free mice (48), resulting in a
single community. Studies examining difference in disease often use
different \emph{C. difficile} strains or ribotypes in mice with similar
microbiota as a proxy for variation in disease, such as strain 630 for
non-severe and RT027 for severe (22, 32, 33, 49). Studies have also
demonstrated variation in infection through tapering antibiotic dosage
(18, 23, 50) or by reducing the amount of \emph{C. difficile} cells or
spores used for the challenge (22, 50). These studies either lack
variation in the initial microbiome or have variation in the \emph{C.
difficile} infection itself, therefore we cannot make any association
between variation in severity and the microbiome.

>>>>

Nicholas A. Lesniak\(^1\), Alyxandria M. Schubert\(^1\), Kaitlyn J.
Flynn\(^1\), Jhansi L. Leslie\(^{1,4}\), Hamide Sinani\(^1\), Ingrid L.
Bergin\(^3\), Vincent B. Young\(^{1,2}\), Patrick D.
Schloss\(^{1,\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}\\
1. Department of Microbiology and Immunology, University of Michigan,
Ann Arbor, MI\\
2. Division of Infectious Diseases, Department of Internal Medicine,
University of Michigan Medical School, Ann Arbor, MI\\
3. Unit for Laboratory Animal Medicine, University of Michigan, Ann
Arbor, MI\\
4. Current affiliation: Department of Medicine, Division of
International Health and Infectious Diseases, University of Virginia
School of Medicine, Charlottesville, Virginia, USA

@@

The severity of \emph{Clostridioides difficile} infections (CDI) has
increased over the last few decades. Patient age, white blood cell
count, creatinine levels as well as \emph{C. difficile} ribotype and
toxin genes have been associated with disease severity. However, it is
unclear whether there is an association between the gut microbiota and
disease severity. The gut microbiota is known to interact with \emph{C.
difficile} during infection. Perturbations to the gut microbiota are
necessary for \emph{C. difficile} to colonize the gut. The gut
microbiota is thought to impair \emph{C. difficile} colonization through
bile acid metabolism, nutrient consumption and bacteriocin production.
Here we sought to demonstrate that members of the gut bacterial
communities can contribute to disease severity. We derived diverse gut
communities by colonizing germ-free mice with different human fecal
communities. The mice were then infected with a single \emph{C.
difficile} ribotype 027 clinical isolate which resulted in moribundity
and histopathologic differences. The variation in severity was
associated with the human fecal community that the mice received.
Generally, bacterial populations with pathogenic potential, such as
\emph{Escherichia}, \emph{Helicobacter}, and \emph{Klebsiella}, were
associated with more severe outcomes. Bacterial groups associated with
fiber degradation, bile acid metabolism and lantibiotic production, such
as \emph{Anaerostipes} and \emph{Coprobacillus}, were associated with
less severe outcomes. These data indicate that gut bacteria can
influence CDI disease severity.

@@

\emph{Clostridioides difficile} colonization can result in no symptoms
or develop into an infection, ranging in severity from mild diarrhea to
toxic megacolon, sepsis and death. Models that predict severity and
guide treatment decisions are based on clinical factors and \emph{C.
difficile} characteristics. Although the gut microbiome plays a role in
protecting against CDI, it has not been investigated for its effect on
CDI disease severity or incorporated into attempts to predict disease
severity. We demonstrated that variation in the microbiome of mice
colonized with human feces yielded a range of disease outcomes. These
results revealed groups of bacteria associated with both severe and mild
\emph{C. difficile} infection outcomes. Gut bacterial community data
from patients with CDI could improve our ability to identify patients at
risk of developing more severe disease and improve interventions which
target \emph{C. difficile} and the gut bacteria to reduce host damage.

@@

\emph{Clostridioides difficile} infections (CDI) have increased in
incidence and severity since \emph{C. difficile} was first identified as
the cause of antibiotic-associated pseudomembranous colitis (1). CDI
disease severity can range from asymptomatic to mild diarrhea to toxic
megacolon and death. The Infectious Diseases Society of America (IDSA)
and Society for Healthcare Epidemiology of America (SHEA) guidelines
defines severe CDI in terms of a white blood cell count greater than
15,000 cells/mm\(^3\) and/or a serum creatinine greater than 1.5 mg/dL.
Patients who develop shock or hypotension, ileus, or toxic megacolon are
considered to have fulminant CDI (2). Since these measures are outcomes
of CDI, these measures have limited ablity to predict risk of severe CDI
when CDI is fiest detected. Schemes have been developed to score a
patient's risk for severe CDI outcomes based on clinical factors but
have not been robust for broad application (3). Thus, we have limited
ability to prevent patients from developing severe CDI. Ideally, we
could identify patients that are high-risk of severe disease so their
treatments could be adjusted to alter the clinical trajectory.

Missing from CDI severity prediction models are the effects of the
indigenous gut bacteria. \emph{C. difficile} interacts with the gut
community in many ways. The indigenous bacteria of a healthy intestinal
community provide a protective barrier preventing \emph{C. difficile}
from infecting the gut. A range of mechanisms can disrupt this barrier,
including antibiotics, medications, or dietary changes, and lead to
increased susceptibility to CDI (4--6). Once \emph{C. difficile}
overcomes the protective barrier and colonizes the intestine, the
indigenous bacteria can either promote or inhibit \emph{C. difficile}
through producing molecules or modifying the environment (7, 8). Bile
acids metabolized by the gut bacteria can inhibit \emph{C. difficile}
growth and affect toxin production (9, 10). Bacteria in the gut also can
compete more directly with \emph{C. difficile} through antibiotic
production or nutrient consumption (11--13). While the relationship
between the gut bacteria and \emph{C. difficile} has been established,
the effect the gut bacteria can have on CDI disease severity is unclear.

Here, we sought to determine whether the gut bacteria contribute to
disease severity. We hypothesized that since the gut bacteria affect the
metabolism of \emph{C. difficile} and its infection dynamics, it can
also affect the disease severity of the infection. To test this
hypothesis, we colonized germ-free C57BL/6 mice with human fecal samples
to create varied gut communities. We then challenged the mice with a
single \emph{C. difficile} isolate, ribotype 027 (RT027) clinical
isolate 431, and followed the mice for the development of severe
outcomes of moribundity and histopathologic cecal tissue damage. Since
the murine host and \emph{C. difficile} isolate were the same and only
the gut community varied, the variation in disease severity we observed
was attributable to the microbiome.

@@

\textbf{\emph{C. difficile} is able to infect germ-free mice colonized
with human fecal microbial communities without antibiotics} To produce
gut microbiomes with greater variation than those found in conventional
mouse colonies, we colonized germ-free mice with bacteria from human
feces (14). We inoculated germ-free C57BL/6 mice with homogenized feces
from one of 15 human fecal samples via oral gavage. These human fecal
samples were selected because they represented diverse community
structures based on community clustering (15). The gut communities were
allowed to equlibriate for two weeks post inoculation (16). We then
surveyed the bacterial members of the gut communities by 16S rRNA gene
sequencing of murine fecal pellets (Figure 1A). The bacterial
communities from each mouse grouped more closely to those communities
from mice that received the same human fecal donor community than to the
mice who received a different human fecal donor community (Figure 1B).
The communities were primarily composed of populations of
\emph{Clostridia}, \emph{Bacteroidia}, \emph{Erysipelotrichia},
\emph{Bacilli}, and \emph{Gammaproteobacteria}. However, the gut
bacterial communities of each donor group of mice harbored unique
relative abundance distributions of the shared bacterial classes.

Next, we tested this set of mice with their human-derived gut microbial
communities for susceptibility to \emph{C. difficile} infection. A
typical mouse model of CDI requires pre-treatment of conventional mice
with antibiotics, such as clindamycin, to become susceptible to \emph{C.
difficile} colonization (17, 18). However, we wanted to avoid modifying
the gut communities with an antibiotic to maintain their unique
microbial compositions and ecological relationships. Since some of these
communities came from people at increased risk of CDI, for example due
to hospitalization or recent antibiotic use (15), we decided to test if
\emph{C. difficile} was able to infect these mice without any antibiotic
perturbation. We hypothesized that \emph{C. difficile} would be able to
colonize the mice who received their gut communities from a donor with
perturbed community. Mice were challenged with 10\(^{3}\) \emph{C.
difficile} RT027 clinical isolate 431 spores. The mice were followed for
10 days post-challenge, and their stool was collected and plated for
\emph{C. difficile} colony forming units (CFU) to determine the extent
of the infection. Surprisingly, communities from all donors were able to
be colonized (Figure 2). Two mice were able to resist \emph{C.
difficile} colonization, both received their community donor N3, which
may be attributed to experimental variation since this group also had
more mice. By colonizing germ-free mice with different human fecal
communities, we were able to generate diverse gut communities in mice,
which were susceptible to \emph{C. difficile} infection without further
modification of the gut community.

\textbf{Infection severity varies by initial community.} After we
challenged the mice with \emph{C. difficile}, we investigated the
outcome from the infection and its relationship to the initial
community. We followed the mice for 10 days post-challenge for
colonization density, toxin production, and mortality. Seven mice, from
Donors N1, N2 and N5, were not colonized at detectable levels on the day
after \emph{C. difficile} challenge but were infected
(\textgreater10\(^{6}\)) by the end of the experiment. All mice that
received their community from Donor M1 through M6 succumbed to the
infection and became moribund within 3 days post-challenge. The
remaining mice, except the uninfected Donor N3 mice, maintained \emph{C.
difficile} infection through the end of the experiment (Figure 2). At 10
days post-challenge, or earlier for the moribund mice, mice were
euthanised and fecal material were assayed for toxin activity and cecal
tissue was collected and scored for histopathologic signs of disease
(Figure 3). Overall, there was greater toxin activity detected in the
stool of the moribund mice (\emph{P} = 0.003). However, when looking at
each group of mice, we observed a range in toxin activity for both the
moribund and non-moribund mice (Figure 3A). Non-moribund mice from
Donors N4 through N9 had comparable toxin activity as the moribund mice.
Additionally, not all moribund mice had toxin activity detected in their
stool. Next we examined the cecal tissue for histopathologic damage.
Moribund mice had high levels of epithelial damage, tissue edema, and
inflammation (Figure S1) similar to previously reported histopathologic
findings for \emph{C. difficile} RT027 (19). As observed with toxin
activity, the moribund mice had higher histopathologic scores than the
non-moribund mice (\emph{P} \textless{} 0.001). However, unlike the
toxin activity, all moribund mice had consistently high histopathologic
summary scores (Figure 3B). The non-moribund mice, Donor groups N1
through N9, had a range in tissue damage from none detected to similar
levels as the moribund mice, which grouped by community donor. Together,
the toxin activity, histopathologic score, and moribundity showed
variation across the donor groups but were largely consistent within
each donor group.

\textbf{Microbial community members explain variation in CDI severity.}
We next interrogated the bacterial communities at the time of \emph{C.
difficile} challenge (day 0) for their relationship to infection
outcomes using linear discriminant analysis (LDA) effect size (LEfSe)
analysis to identify individual bacterial populations that could explain
the variation in disease severity. We split the mice into groups by
severity level based on their moribundity and histopathologic score. We
dichotomized the histopathologic scores into high and low groups by
splitting on the median score of 5. This analysis revealed 20 genera
that were significantly different by the disease severity (Figure 4A).
Bacterial genera \emph{Turicibacter}, \emph{Streptococcus},
\emph{Staphylococcus}, \emph{Pseudomonas}, \emph{Phocaeicola},
\emph{Parabacteroides}, \emph{Bacteroides}, and
\emph{Escherichia/Shigella} were detected at higher relative abundances
in the mice that became moribund. Populations of \emph{Anaerotignum},
\emph{Coprobacillus}, \emph{Enterocloster}, and \emph{Murimonas} were
more abundant in the non-moribund mice that would develop only low
intestinal injury. To understand the role of toxin activity in disease
severity, we applied LEfSe to identify the genera most likely to explain
the differences between the presence and absence of detected toxin
activity (Figure 4B). Many genera that associated with toxin production
were also associated with moribundity, such as populations of
\emph{Escherichia/Shigella} and \emph{Bacteroides}. Likewise, there were
genera such as \emph{Anaerotignum}, \emph{Enterocloster}, and
\emph{Murimonas} that were associated with no detected toxin production
that also exhibited greater relative abundance in communities from
non-moribund mice with a low histopathologic score. Lastly, we tested
for correlations between the endpoint relative abundances of bacterial
operational taxonomic units (OTUs) and the histopathologic summary score
(Figure 4C). The endpoint relative abundance of \emph{Bacteroides} was
positively correlated with histopathologic score, as its day 0 relative
abundance did with disease severity (Figure 4A). Populations of
\emph{Klebsiella} and \emph{Prevotellaceae} were positively correlated
with the histopathologic score and were increased in the group of mice
with detectable toxin. This analysis identified bacterial genera that
were associated with the variation in moribundity, histopathologic
score, and toxin activity of CDI.

@@

@@

Challenging mice colonized with different human fecal communities with a
single \emph{C. difficile} isolate, RT027 isolate 431, demonstrated that
variation in the gut bacteria affects the \emph{C. difficile} infection
disease severity. Our analysis revealed an association between the
relative abundance of bacterial community members and disease severity.
Previous studies investigating the severity of CDI disease involving the
microbiome have had limited ability to interrogate this relationship
between the microbiome and disease severity. Studies that have used
clinical data have limited ability to control variation in the host,
microbiome or \emph{C. difficile} ribotype (20). Murine experiments
typically use a single mouse colony and different \emph{C. difficile}
ribotypes to create severity differences (21). Recently, our group has
begun uncovering the effect microbiome variation has on \emph{C.
difficile} infection. We showed the variation in the bacterial
communities between mice from different mouse colonies resulted in
different clearance rates of \emph{C. difficile} (22). We also showed
varied ability of mice to spontaneously eliminate \emph{C. difficile}
infection when they were treated with different antibiotics prior to
\emph{C. difficile} challenge (23). Overall, the results presented here
have demonstrated that the gut bacterial community contributed to the
severity of \emph{C. difficile} infection.

\emph{C. difficile} can lead to asymptomatic colonization or infections
with severity ranging from mild diarrhea to death. Physicians use
classification tools to identify patients most at risk of developing a
severe infection using white blood cell counts, serum albumin level, or
serum creatinine level (2, 24, 25). Those levels are driven by the
activities in the intestine (26). Research into the drivers of this
variation have revealed factors that make \emph{C. difficile} more
virulent. Strains are categorized for their virulence by the presence
and production of the toxins TcdA, TcdB, and binary toxin and the
prevalence in outbreaks, such as ribotypes 027 and 078 (17, 27--30).
However, other studies have shown that disease is not necessarily linked
with toxin production (31) or the strain (32). Furthermore, there is
variation in the genome, growth rate, sporulation, germination, and
toxin production in different isolates of a strain (33--36). This
variation may help explain why severe CDI prediction tools often miss
identifying many patients with CDI that will develop severe disease (3,
21, 37, 38). Therefore, it is necessary to gain a full understanding of
all factors contributing to disease variation to improve our ability to
predict severity.

The state of the gut bacterial community determines the ability of
\emph{C. difficile} to colonize and persist in the intestine. \emph{C.
difficile} is unable to colonize an unperturbed healthy gut community
and is only able to become established after a perturbation (18). Once
colonized, the different communities lead to different metabolic
responses and dynamics of the \emph{C. difficile} population (8, 23,
39). Gut bacteria metabolize primary bile acids into secondary bile
acids (40, 41). The concentration of these bile acids affects
germination, growth, toxin production and biofilm formation (9, 10, 42,
43). Members of the bacterial community also affect other metabolites
\emph{C. difficile} utilizes. \emph{Bacteroides thetaiotaomicron}
produce sialidases which release sialic acid from the mucosa for
\emph{C. difficile} to utilize (44, 45). The nutrient environment
affects toxin production (46). Thus, many of the actions of the gut
bacteria modulate \emph{C. difficile} in ways that could affect the
infection and resultant disease.

A myriad of studies have explored the relationship between the
microbiome and CDI disease. CDI studies often use an infection model
with a single homogeneous bacterial community. Collins \emph{et al.}
used multiple human communities to colonize mice, however the
communities were pooled prior to gavaging into germ-free mice (47),
resulting in a single human-derived community. Studies examining
difference in disease often use different \emph{C. difficile} strains or
ribotypes in mice with similar microbiota as a proxy for variation in
disease, such as strain 630 for non-severe and RT027 for severe (17, 27,
28, 48). Studies have also demonstrated variation in infection through
tapering antibiotic dosage (18, 23, 49) or by reducing the amount of
\emph{C. difficile} cells or spores used for the challenge (17, 49).
These studies either lack variation in the initial microbiome or have
variation in the \emph{C. difficile} infection itself, confounding any
association between variation in severity and the microbiome.
\textbf{Add Hikiro et al} With our recent observations that the initial
community affected the ability of \emph{C. difficile} to persist in the
gut (22, 23) and the existing research describing the numerous
interactions between the microbiome and \emph{C. difficile}, we
hypothesized that the microbiome can modulate CDI disease severity. Our
data have demonstrated gut bacterial relative abundances associate with
variation in toxin production, histopathologic scoring of the cecal
tissue and mortality. This analysis revealed populations of
\emph{Akkermansia}, \emph{Anaerostipes}, \emph{Coprobacillus},
\emph{Enterocloster}, \emph{Lactonifactor}, and \emph{Monoglobus} were
more abundant in the microbiome of non-moribund mice which had low
histopathologic scores and no detected toxin. The protective role of
these genera are supported by previous studies. \emph{Coprobacillus},
\emph{Lactonifactor}, and \emph{Monoglobus} have been shown to be
involved in dietary fiber fermentation and associated with healthy
communities (50--53). \emph{Anaerostipes} and \emph{Coprobacillus},
which produce short chain fatty acids, have been associated with healthy
communities (54--56). Furthermore, \emph{Coprobacillus}, which was
abundant in mice with low histopathologic scores but rare in all other
mice, has been shown to contain a putative type I lantibiotic gene
cluster and inhibit \emph{C. difficile} colonization (57--59).
\emph{Akkermansia} and \emph{Enterocloster} were also identified as more
abundant in mice which had a low histopathologic scores but have
contradictory supporting evidence in the current literature. In our
data, \emph{Akkermansia} was most abundant in the non-moribund mice with
low histopathologic scores but there were some moribund mice which had
increased populations of \emph{Akkermansia}. This could be attributed to
either a more protective mucus layer was present inhibiting colonization
(59, 60) or mucus consumption by \emph{Akkermansia} could have been
crossfeeding \emph{C. difficile} or exposing a niche for \emph{C.
difficile} (61--63). Similarly, \emph{Enterocloster} was more abundant
and associated with low histopathologic scores. It has been associated
with healthy populations and has been used to mono-colonize germ-free
mice to reduce the ability of \emph{C. difficile} to colonize (64, 65).
However, \emph{Enterocloster} has also been involved in infections, such
as bacteremia (66, 67). These data have exemplified populations of
bacteria that have the potential to be either protective or harmful.
Thus, the disease outcome is not likely based on the abundance of
individual populations of bacteria, rather it is the result of the
interactions of the community.
